Temporal Continuity and the Judgment of Actual Causation

### 2021-05-04 docs/resources/CogSci_paper/paper.tex
explain prob raising prediction

\documentclass[10pt,letterpaper]{article}
\usepackage{cogsci}
%\usepackage{dsfont}
\usepackage{bm}
\usepackage{amssymb}
\usepackage{indentfirst}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{multirow}

%\cogscifinalcopy % Uncomment this line for the final submission

@@

\newcommand{\ev}[2]{$#1_#2$}
\renewcommand\bibliographytypesize{\footnotesize}

@@

\author{{\large \bf Morton Ann Gernsbacher (MAG@Macc.Wisc.Edu)} \\
  Department of Psychology, 1202 W. Johnson Street \\
  Madison, WI 53706 USA
  \AND {\large \bf Sharon J.~Derry (SDJ@Macc.Wisc.Edu)} \\
  Department of Educational Psychology, 1025 W. Johnson Street \\
  Madison, WI 53706 USA}

@@

Identifying the cause of an event (e.g.\ the stopping of an elevator) requires a judgment of actual causation, also known as singular or token causation~\cite{danks17}. Psychologists and philosophers have explored these judgments in detail~and have developed formal models of actual causation, including models based on Bayesian networks \cite{halpern16}, force dynamics \cite{wolfft17} and mental simulation~\cite{gerstenberg20}.  Here we present and evaluate an account of actual causation that highlights the role of temporal continuity.  On this account, the tripping of relay Q29BD is the main cause of the elevator stopping because the tripping initiated a continuous sequence of state changes that culminated in the stopping of the elevator.

@@

The results of Experiment 1 are broadly consistent with both the continuity and probability raising accounts, and the goal of Experiment 2 was to distinguish between these accounts. The two make different predictions for dual branch sequences in which the delayed branch activates after activation has already begun along the continuous branch. One such sequence is shown in Figure~\ref{fig:intro}b. 
In cases like this, \ev{R}{D} guarantees the occurrence of the effect, and therefore qualifies as the primary cause according to the probability raising account. The continuity account, however, still treats \ev{R}{C} as the primary cause.

The materials and procedure for Experiment 2 are similar to those for Experiment 1, and we highlight only the few points of difference.

@@

@@

\bibliography{uber}

>>>>

\documentclass[10pt,letterpaper]{article}
\usepackage{cogsci}
%\usepackage{dsfont}
\usepackage{bm}
\usepackage{amssymb}
\usepackage{indentfirst}
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{balance}

\cogscifinalcopy % Uncomment this line for the final submission

@@

\newcommand{\ev}[2]{$#1_#2$}
%\renewcommand\bibliographytypesize{\footnotesize}

@@

\author{{\large \bf Aur\'{e}lien Fermo (XXX)} \\
  XXX \\
  XXX 
  \AND {\large \bf Charles Kemp (c.kemp@unimelb.edu.au)} \\
  School of Psychological Sciences, University of Melbourne\\
  Parkville, Victoria 3010, Australia}

@@

Identifying the cause of an event (e.g.\ the stopping of an elevator) requires a judgment of actual causation, also known as singular or token causation~\cite{danks17}. Psychologists and philosophers have explored these judgments in detail~and have developed formal models of actual causation, including models based on Bayesian networks \cite{halpern16}, force dynamics \cite{wolfft17} and mental simulation~\cite{gerstenberg21}.  Here we present and evaluate an account of actual causation that highlights the role of temporal continuity.  On this account, the tripping of relay Q29BD is the main cause of the elevator stopping because the tripping initiated a continuous sequence of state changes that culminated in the stopping of the elevator.

@@

The results of Experiment 1 are broadly consistent with both the continuity and probability raising accounts, and the goal of Experiment 2 was to distinguish between these accounts. The two make different predictions for dual branch sequences in which the delayed branch activates after activation has already begun along the continuous branch. One such sequence is shown in Figure~\ref{fig:intro}b. 
In cases like this, the continuity account still treats \ev{R}{C} as the primary cause, but the probability raising account now treats \ev{R}{D} as the primary cause. To see why, let $b$ (for baserate) be the probability of activation along along a given branch during a given trial. At the start of a dual branch trial, the probability that the GCM will activate is $b^2$, because independent activations are required along both branches. After \ev{R}{C} the probability increases to $b$, and after \ev{R}{D} the probability increases to 1. The two probability increments are therefore $b - b^2$ and $1 - b$, and it is straightforward to show that $b - b^2 \le 1 - b$ for all values of $b$, with equality attained only when $b = 1$.  As a result, the probability raising account chooses \ev{R}{D} for virtually all values of the baserate, and if $b=1$ the account is indifferent between \ev{R}{C} and \ev{R}{D}.

The materials and procedure for Experiment 2 are similar to those for Experiment 1, and we will highlight only the few points of difference.  As for Experiment 1, no explicit information about the baserate $b$ was provided. For each network shown, activation eventually occurred along each branch, and some participants may have therefore assumed that $b=1$. Most participants, however, probably implicitly assumed that $b < 1$, which allows for the possibility of observation periods during which no activations occurred. We therefore assume that $b < 1$ when discussing the predictions of the probability raising account.

@@

\section{Acknowledgments}

AF carried out this work at the University of Melbourne while enrolled in a Master's degree at the \'{E}cole Normale Sup\'{e}rieure. 
% add financial support?

@@

\balance
\bibliography{uber}
